Trauma treatment delay

### 2023-09-21 manuscriptlongversion.Rmd
intro

Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, it has been shown that this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. But it is not only patient level factors that affect what effect trauma has on people. About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma. Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

**Mortality and morbidity conferences**

Many hospitals, mainly in HIC, have a trauma registry where they log the patients information and timeline of what happened. This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions. OFIs are the endpoint of MoM conferences. Once

This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database. These MoM conferences are very resource and time craving, which is why many hospitals doesn't<!-- Do not use contractions--> have the means to perform them. In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived. Examples of opportunities for improvement may include lack of resources and management errors.

>>>>

Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries (HIC), depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, it has been shown that this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. But it is not only patient level factors that affect what effect trauma has on people. About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma. Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

**Mortality and morbidity (MoM) conferences**

Many hospitals, mainly in HIC, have a trauma registry where they log the patients information and timeline of what happened. This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences. The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors. These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].

OFIs are the endpoint of MoM conferences. These meetings are

where possible opportunities for improvement (OFI) were identified by nurses who go through the trauma care quality database

These MoM conferences are very resource and time craving, which is why it is a concept that has not been broadly implemented.

In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived. Examples of opportunities for improvement may include lack of resources and management errors.
